Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-04-24 paper.tex
Updates from Yan

\subsection{Imputation of Pango lineage for non-sample nodes}
Non-sample nodes inserted during ARG inference had no Pango lineage
assignment from Nextclade. We imputed the Pango lineage of non-sample nodes
using a two-step procedure (Figure 3) to help assess how consistently
breakpoint intervals are estimated by the HMM runs (see below). If the Pango
lineage of the parental node was available and no mutation was immediately
ancestral to the non-sample node, then the parental node’s status was directly
propagated to the non-sample node. Otherwise, the Pango lineage of the
non-sample node was inferred by matching its full set of mutations against the
Pango lineage-defining mutations (based on 90\% consensus of the sequences
analysed) from the COVID-CG website \citep{Chen2021-zc};
\url{https://covidcg.org/}; accessed on November 04, 2022).

@@

\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/recombination_node_mrcas.pdf}
\caption{\label{fig:recomb_mrcas} Divergance between parents at recombination breakpoints. (A) All recombination nodes that are ancestral to least two samples. (B) Further filtered as described in the text.
MORE DETAIL. Horizontal lines represent WHAT,
Figure~\ref{fig:recomb_mrcas_voc_breakdown} shows the same data broken down by parental ``VoC'' classification.}
\end{figure}

@@

\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/supp_recombination_node_mrcas.pdf}
\caption{\label{fig:recomb_mrcas_voc_breakdown}  DESCRIBE ME}
\end{figure}

>>>>

\subsection{Imputation of Pango lineage for non-sample nodes}
Initially, only sample nodes have Pango lineage assignments.
For analysis of recombinant parents, it is also helpful to assign
Pango lineages to non-sample nodes. We do this
using a two-step procedure (Figure 3). If the Pango
lineage of the parental node was available and no mutation was immediately
ancestral to the non-sample node, then the parental node’s status was directly
propagated to the non-sample node. Otherwise, the Pango lineage of the
non-sample node was inferred by matching its full set of mutations against the
Pango lineage-defining mutations (based on 90\% consensus of the sequences
analysed) from the COVID-CG website \citep{Chen2021-zc};
\url{https://covidcg.org/}; accessed on November 04, 2022).

@@

\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/recombination_node_mrcas.pdf}
\caption{\label{fig:recomb_mrcas} Date of common ancestry between the parents either side of recombination breakpoints, as a function of degree of parental divergence. MRCAs of parents inferred to have produced a Pango designated recombinant (XA, XB, etc.) are identified in orange (A) Data after removing singleton recombinants (XXX parent pairs). Many recombinations involve parental strains that descend from the same MRCA: dates for the four most frequently identified MRCA nodes are indicated by horizontal dotted lines. These frequent MRCA nodes tend to be associated with major waves and commonly have many immediate children (i.e. are large polytomies, likely to indicate undersampled outbreaks) (B) Further filtered by removing HMM inconsistent breakpoints (YYY parent pairs). Note that Figure~\ref{fig:recomb_mrcas_voc_breakdown} shows the same data broken down by parental VoC classification.}
\end{figure}

@@

\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/supp_recombination_node_mrcas.pdf}
\caption{\label{fig:recomb_mrcas_voc_breakdown}  Divergence between parent lineages for recombination events within and among different VoC categories. HMM-consistent recombinants are shown in green, HMM-inconsistent in blue (including many of the named Delta-Omicron recombinants). We find 69 HMM consistent (77 inconsistent) Alpha-Alpha recombination breakpoints = 13\% (10\%) of the total; 84 HMM consistent (143 inconsistent) Delta-Delta recombination breakpoints = 18\% (18\%) of the total; 99 HMM consistent (155 inconsistent) Omicron-Omicron recombination breakpoints = 22\% (19\%) of the total; and 2 HMM consistent (20 inconsistent) Delta-Omicron recombination breakpoints = 2\% (< 1\%) of the total. These plots do not include recombination breakpoints involved lineages that were not classified into these VoC categories}
\end{figure}
